A spatial analysis of the environmental correlates of COVID-19 incidence in Spain

### 2020-04-18 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.tex
Edits to Context and Methods

For a better interpretation of the results of our analysis is necessary
to know the chronology of the epidemic outbreak in Spain. This
subsection presents a brief time line of COVID-19.

The first reported case of COVID-19 in Spain was on January 31th, were a
German tourist in the Canary Islands was positively tested, but the
first not imported case was diagnostic on February 27th in Sevilla
province (Andalusia). The epidemic was growing slowly at the beginning
but in March 11th the World Health Organization (WHO) declared
officially the pandemic. This declaration was a break point and on March
13th the Spanish National Government declare the ``Emergency Status''.
At this date, Spain have been declared 4209 cases of COVID-19 (mostly of
cases, 1990, in Madrid). Only essential services (e.g.~food, health) and
some economic subsector of industry and construction could to maintain
activity. On March 17th Spain closes its lands borders and only allows
entry to nationals and residents. From March 30th to April 12th (April
10th-12th Easter) lockdown of the Spanish people was hardened and only
essential activities were allowed.

@@

Our dataset includes information about the daily number of cases of
COVID-19 reported in Spain at the provincial level (NUTIII in Eurostat
terminology) for the period between March 13th and April 11th,
inclusive. This gives a temporal coverage of 30 days and spatial
coverage of 50 units. For our purposes, we consider positive cases
reported, but excluding symptomatic cases diagnosed by a doctor without
a Polymerase Chain Reaction (PCR) test. The Spanish National Government
publishes periodic updates at the regional level (NUTII) and the
information is also released at the provincial level as part of a
collaborative project {[}\textbf{by whom?}{]}. This information is
compiled from several sources, mainly the official web pages of the
Spanish regional goverments. In addition, we consider two sets of
explanatory variables. The first one, and the focus of this research, is
set of two environmental variables, namely temperature and humidity,
which are collected from official sources (i.g., AEMET, the state
meteorology agency, and MAPA, the ministry of agriculture, fisheries,
and food). The second set provides some relevant controls for
multivariate analysis, and refers to economic and demographic attributes
of the province (also collected from official sources, i.e., INE, the
national statistics institute). Table \ref{tab:descriptive-statistics}
shows the descriptive statistics and the provenance of the data.

@@

\caption{\label{tab:descriptive-statistics}\label{tab:descriptive-statistics}Descriptive statistics}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{l>{\raggedright\arraybackslash}p{15em}rrrr>{\raggedright\arraybackslash}p{5em}}
\toprule
Variable & Note & Min & Mean & Max & SD & Source\\
\midrule
\rowcolor{gray!6}  COVID-19 Incidence & Incidence in reported cases of SARS-19 per 100,000 people & 0.38 & 153.61 & 1149.36 & 186.23 & Montera34\\
GDPpc & GDP per capita in €1,000s & 16.67 & 22.51 & 36.00 & 4.77 & INE\\
\rowcolor{gray!6}  Older & Percentage of people aged 65 and older in the province & 15.16 & 21.03 & 31.36 & 3.95 & INE\\
Population Density & Population density in the province in people per sq.km & 8.60 & 140.04 & 829.76 & 181.25 & INE\\
\rowcolor{gray!6}  Mean Temperature & Mean temperature in province by date, in C & 1.00 & 12.18 & 23.20 & 3.67 & AEMET\\
\addlinespace
Humidity & Relative humidity in province by date & 2.00 & 77.82 & 100.00 & 10.37 & MAPA\\
\bottomrule
\multicolumn{7}{l}{\textit{Note: }}\\
\multicolumn{7}{l}{Montera34: https://code.montera34.com:4443/numeroteca/covid19/-/blob/master/README\_providencialdata19.md}\\
\multicolumn{7}{l}{INE (Instituto Nacional de Estadistica): https://www.ine.es/}\\
\multicolumn{7}{l}{AEMET (Agencia Estatal de Meteorologia): http://eportal.mapa.gob.es}\\
\multicolumn{7}{l}{MAPA (Ministerio de Agricultura, Pesca y Alimentacion): http://eportal.mapa.gob.es}\\
\end{tabular}}
\end{table}

@@

The Seemingly Unrelated Regression equations model (SUR from now on) is
a multivariate econometric formulation used in different fields when the
data set is organized as cross-sections for different time periods. The
basis of this approach is well-known since the initial works of Zellner
(1962) and now is a popular methodology included in several Econometrics
textbook (e.g., Greene, 2003). To our knowledge, Anselin (1988) was the
first text that introduced the term spatial SUR in reference to a model
made of `an equation for each time period, which is estimated for a
cross section of spatial units' (p.~141). From this milestone, a huge
researches used this methodology in spatial framework. Rey and Montouri
(1999), Fingleton ({\textbf{???}}), Lauridsen et al.~(2010), Le Gallo
and Dall'Erba (2006), and Lopez et al.~(2017) are well-known examples in
this line.

The classical SUR model without spatial effects (from here, SUR-SIM) is
a stack of equations as follows:

@@

\noindent where \(y_{t}=(y_{1t},...,y_{Nt})\) is a \(N \times 1\)
vector, and in our case \(y_{st}\) is the incidence ratio in the
province \(s\) (\(s=1,...,N\)) the day \(t\) \((t=1,...,T)\);
\(X_t=(X^1,...,X^{k_t})\) is a \(N \times k_t\) matrix of the \(k_t\)
independent variables, \(X_i=(X^i_{st})\);
\(\beta_t=(\beta_{1t},...,\beta_{Nt})\) is a vector of coefficients and
\(\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})\) is the vector of
residuals.

The more relevant characteristic of a SUR model is the dependence
structure among the vectors of residuals, namely:

@@

Note that this specification is very flexible, in that it allows changes
in the coefficients \(\beta_{it}\) in order to modulate the effect of
\(X^i_{.t}\) on \(y_t\). This flexibility can be reduced and it is
posible to impose restrinctions considering some \(\beta\) coefficients
as being constant over time. In this case, we can reformulate the
coefficients expression of
\(\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})\)
considering the first \(r\) coefficients contanst.

Equation \ref{eq:sur-sim}) can be rewriten in compact form:

@@

As is the case of cross-sectional data, it is possible identify spatial
autocorrelation in the residuals of Model \ref{eq:sur-sim}, and several
Lagrange Multipliers tests have been developed to test the null
hypothesis of spatial independence (see López et al., 2014). When the
null hypothesis is rejected, there are alternative several
specifications that have been propose to include spatial effects
(Anselin, 1988, see also 2016). In this paper we consider a spatial SUR
model that incorporates a spatial lag of the dependent variable as an
explanatory factor. The stack expresion for the spatial lag SUR model
(SUR-SLM),

@@

\noindent where A =\(I_{TN}-\bf{\Lambda} \otimes W\) with
\(\bf{\Lambda} = diag(\lambda_1, \cdots, \lambda_T)\).

This specification assumes that incidence in a province (\(y_st\)) at
time \(t\) is partially determined by the weighted average (\(Wy_{st}\))
of incidence in neighbouring provinces. Parameters of spatial dependence
are estimated for each time period \(\lambda_t\) and identifies the
intensity and the sign of the impacts of neighbourhood. It is possible
test the null hypotheis of identtical levels of spatial dependence
(\(\lambda_i=\lambda_j, \forall i,j\)). The correspond Wald test is
available in the R package \textbf{spsur}

@@

Figure \ref{fig:weekly-average-incidence-map} shows the geographical
variation in the incidence of the disease, as well as its temporal
progression in weekly averages. Our analysis begins on March 13, which
is the first date when every province had reported at least one case of
COVID-19. It can be seen that the highest incidence at this early date
was in the provice of Álava, in the North of Spain. While not the most
populous province, with a population of only 331,549, Álava has the
highest GDP per capita of all provinces. Vitoria, its main city, is the
capital of the Basque country and has been the focus of efforts to
develop a ``Global Basque City'' (Meijers et al., 2008), along with San
Sebastian and Bilbao. The other early focus of the pandemic in Spain was
Madrid, which is the most populous province in the country and has the
second highest GDP per capita after Álava. The early outbreaks in these
two provinces can be traced throughout the progression of the pandemic
over time, although by the end of the period under consideration, other
provinces had matched and even surpased their incidence rates, including
Segovia and Soria to the north of Madrid, and Ciudad Real and Albacete
to the south.

@@

>>>>

The first reported case of COVID-19 in Spain was on January 31th, when a
German tourist in the Canary Islands tested positive for the virus.
However, it was still a few weeks before the first domestic case was
reported, on February 27th in Sevilla province (Andalusia). In a short
period of time, after this relatively slow start, the outbreak flared.
By March 11th the World Health Organization (WHO) declared COVID-19
officially a pandemic. This declaration marked a turning point for the
public in Spain too. As of March 13th, the number of cases of COVID-19
reported in Spain was 4,473, with a majority of cases (1,990)
concentrated in Madrid: these numbers were at the time the worst
outbreak in Europe, after Italy. In response to the situation, on March
13th the Spanish National Government declared a state of emergency. As
part of the state of emergency restrictions to most activities were
imposed, with the exception of essential services (e.g.~food, health)
and some economic subsectors of industry and construction. A few days
later, on March 17th, Spain closed its lands borders to allow entry only
to returnee nationals and permanent residents. The lockdown was further
hardened between March 30th and April 12th (including the Easter weekend
of April 10th-12th) and during this period only essential activities
were allowed. During this period, there was a dramatic reduction in
overall mobility, both within provinces
\footnote{https://www.mitma.gob.es/ministerio/covid-19/evolucion-movilidad-big-data/movilidad-provincial}
as between
\footnote{https://www.mitma.gob.es/ministerio/covid-19/evolucion-movilidad-big-data/movilidad-provincial}.

@@

Our dataset includes information about the daily number of cases of
COVID-19 reported in Spain at the provincial level (NUTIII in Eurostat
terminology) for the period between March 13th and April 11th,
inclusive. For our purposes, we consider positive cases reported, but
excluding symptomatic cases diagnosed by a doctor without a Polymerase
Chain Reaction (PCR) test. The Spanish National Government publishes
periodic updates at the regional level (NUTII) and the information is
also released at the provincial level as part of a collaborative project
{[}\textbf{by whom?}{]}. This information is compiled from several
sources, mainly the official web pages of the Spanish regional
goverments. In addition, we consider two sets of explanatory variables.
The first one, and the focus of this research, is set of two
environmental variables, namely temperature and humidity, which are
collected from official sources (i.g., AEMET, the state meteorology
agency, and MAPA, the ministry of agriculture, fisheries, and food). The
second set provides some relevant controls for multivariate analysis,
and refers to economic and demographic attributes of the province (also
collected from official sources, i.e., INE, the national statistics
institute). Table \ref{tab:descriptive-statistics} shows the descriptive
statistics and the provenance of the data.

The spatial and temporal coverage of the data is as follows. Our dataset
begins on March 13, which is the first date when every province had
reported at least one case of COVID-19, and continues until April 11,
for a period of 30 days. The unit of analysis is the province. Provinces
are the equivalent of states, and are embedded in Autonomous
Communities. As an example, Cataluña is an Autonomous Community and
consists of four provinces, namely Barcelona, Gerona, Lerida, and
Tarragona. The size of the provinces is relatively large, as seen in
Table \ref{tab:descriptive-statistics}. The smallest province is
\(1,978.12km^2\) (this is smaller than Rhode Island in the US) and the
largest province is \(21,767.93km^2\) (slightly smaller than New Jersey
in the US). While this is a relatively large degree of spatial
aggregation, reporting on COVID-19 is inconsistent at smaller
geographies, or cases are not reported at that level at all. The
analysis must therefore be considered ecological.

@@

\caption{\label{tab:descriptive-statistics}\label{tab:descriptive-statistics}Descriptive statistics}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{l>{\raggedright\arraybackslash}p{15em}rrrr>{\raggedright\arraybackslash}p{5em}}
\toprule
Variable & Note & Min & Mean & Max & SD & Source\\
\midrule
\rowcolor{gray!6}  COVID-19 Incidence & Incidence in reported cases of SARS-19 per 100,000 people & 0.38 & 153.61 & 1149.36 & 186.23 & Montera34\\
Area & Area of province in sq.km & 1978.12 & 10118.79 & 21767.93 & 4805.98 & INE\\
\rowcolor{gray!6}  GDPpc & GDP per capita in €1,000s & 16.67 & 22.51 & 36.00 & 4.77 & INE\\
Older & Percentage of people aged 65 and older in the province & 15.16 & 21.03 & 31.36 & 3.95 & INE\\
\rowcolor{gray!6}  Population Density & Population density in the province in people per sq.km & 8.60 & 140.04 & 829.76 & 181.25 & INE\\
\addlinespace
Mean Temperature & Mean temperature in province by date, in C & 1.00 & 12.18 & 23.20 & 3.67 & AEMET\\
\rowcolor{gray!6}  Humidity & Relative humidity in province by date & 2.00 & 77.82 & 100.00 & 10.37 & MAPA\\
\bottomrule
\multicolumn{7}{l}{\textit{Note: }}\\
\multicolumn{7}{l}{Montera34: https://code.montera34.com:4443/numeroteca/covid19/-/blob/master/README\_providencialdata19.md}\\
\multicolumn{7}{l}{INE (Instituto Nacional de Estadistica): https://www.ine.es/}\\
\multicolumn{7}{l}{AEMET (Agencia Estatal de Meteorologia): http://eportal.mapa.gob.es}\\
\multicolumn{7}{l}{MAPA (Ministerio de Agricultura, Pesca y Alimentacion): http://eportal.mapa.gob.es}\\
\end{tabular}}
\end{table}

An important aspect of working with environmental data such as
temperature and humidity is the incubation period of the disease. Lauer
et al.~(2020) report for the case of COVID-19 a median incubation period
of 5.7 days (with a confidence interval between 4.9 to 7.8 days). The
vast majority of cases (95\%) develop symptoms between 2.6 days (CI, 2.1
to 3.7 days) and 12.5 days (CI, 8.2 to 17.7 days). For this reason, we
judge it best to use lagged values of the environmental variables. We
test different time lags as follows. We consider a lagged 8-day average,
from date-minus-12 to date-minus-5 days (hereafter \emph{lag8}).
Secondly, we consider a lagged 11-day average, from date-minus-12 to
date-minus-2 days (hereafter \emph{lag11}). Finally, to account for the
likely duration of incubation, we consider a lagged 11-day
\emph{weighted} average, from date-minus-12 to date-minus-2 days
(hereafter \emph{lag11w}). The weights for this variable are calculated
using the parameters of the log-normal distribution reported by Lauer et
al.~(2020), i.e., a log-mean of 1.621 and a log-standard deviation of
0.418. With these weights, the environmental variables at date-minus-2
and date-minus-12 days are weighted as 0.041 and 0.009 respectively,
whereas the environmental variables at date-minus-5 days are weighted as
0.194.

@@

The Seemingly Unrelated Regression equations model (SUR hereafter) is a
multivariate econometric model used in different fields when the
structure of the data consisits of cross-sections for different time
periods. The basis of this approach is well-known since the initial
works of Zellner (1962), and has become a popular methodology included
in several econometrics textbook (e.g., Greene, 2003). To our knowledge,
Anselin (1988) was the first author to discuss SUR from a spatial
perspective. In his landmark text, Anselin discussed a model made of
``an equation for each time period, which is estimated for a cross
section of spatial units'' (p.~141). From this milestone, a large body
of research has developed to extend the classical SUR into a spatial
framework (e.g., Rey and Montouri, 1999; Lauridsen et al., 2010; Le
Gallo and Dall'Erba, 2006; López et al., 2017).

The classical SUR model without spatial effects (hereafter, SUR-SIM) is
a stack of equations as follows:

@@

\noindent where \(y_{t}=(y_{1t},...,y_{Nt})\) is an \(N \times 1\)
vector, and in our case \(y_{st}\) is the incidence ratio in province
\(s\text{ }s=1,...,N\)) on day \(t\text{ }(t=1,...,T)\);
\(X_t=(X^1,...,X^{k_t})\) is an \(N \times k_t\) matrix of the \(k_t\)
independent variables, \(X_i=(X^i_{st})\);
\(\beta_t=(\beta_{1t},...,\beta_{Nt})\) is a vector of coefficients and
\(\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})\) is the usual vector of
random residuals.

A key feature of the SUR framework is the dependence structure among the
vectors of random residuals, which takes the following form:

@@

Note that this specification is very flexible, since it allows the
coefficients \(\beta_{it}\) to vary in order to modulate the effect of
\(X^i_{.t}\) on \(y_t\). A less flexible if more parsimonious model can
be obtained by imposing restrictions to some \(\beta\) coefficients to
fix them across equations. This is equivalent to specifying an effect
that is constant in time. In this case, we can reformulate the
coefficients expression of
\(\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})\)
to restrict the first \(r\) coefficients to be constant.

Equation \ref{eq:sur-sim}) can be rewriten in compact form as follows:

@@

As is the case with cross-sectional data, it is possible to test the
residuals of Model \ref{eq:sur-sim} for spatial autocorrelation, and
several tests have been developed to test the null hypothesis of spatial
independence (see López et al., 2014). When the null hypothesis is
rejected, several alternative specifications have been proposed to
include spatial effects (Anselin, 1988, see also 2016). In this paper we
consider a spatial SUR model that incorporates a spatial lag of the
dependent variable as an explanatory factor. A sp

The stack expresion for the spatial lag SUR model (SUR-SLM),

@@

\noindent where \(A =I_{TN}-\bf{\rho} \otimes W\) with
\(\bf{\rho} = diag(\rho_1, \cdots, \rho_T)\).

This specification assumes that incidence in a province (\(y_st\)) at
time \(t\) is partially determined by the weighted average (\(Wy_{st}\))
of incidence in neighbouring provinces. Coefficients for the spatially
lagged variable are estimated for each time period \(\rho_t\) and
identifies the intensity and the sign of the impacts of neighbourhood.
It is possible test the null hypotheis of identtical levels of spatial
dependence (\(\rho_i=\rho_j, \forall i,j\)). The correspond Wald test is
available in the R package \texttt{spsur}.

@@

Figure \ref{fig:weekly-average-incidence-map} shows the geographical
variation in the incidence of the disease, as well as its temporal
progression in weekly averages. It can be seen that the highest
incidence at this early date was in the provice of Álava, in the North
of Spain. While not the most populous province, with a population of
only 331,549, Álava has the highest GDP per capita of all provinces.
Vitoria, its main city, is the capital of the Basque country and has
been the focus of efforts to develop a ``Global Basque City'' (Meijers
et al., 2008), along with San Sebastian and Bilbao. The other early
focus of the pandemic in Spain was Madrid, which is the most populous
province in the country and has the second highest GDP per capita after
Álava. The early outbreaks in these two provinces can be traced
throughout the progression of the pandemic over time, although by the
end of the period under consideration, other provinces had matched and
even surpased their incidence rates, including Segovia and Soria to the
north of Madrid, and Ciudad Real and Albacete to the south.

@@

\leavevmode\hypertarget{ref-Lauer2020incubation}{}%
Lauer, S.A., Grantz, K.H., Bi, Q., Jones, F.K., Zheng, Q., Meredith,
H.R., Azman, A.S., Reich, N.G., Lessler, J., 2020. The incubation period
of coronavirus disease 2019 (covid-19) from publicly reported confirmed
cases: Estimation and application. Annals of Internal Medicine.
doi:\href{https://doi.org/10.7326/m20-0504}{10.7326/m20-0504}
